---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# extras

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Travis build status](https://travis-ci.com/poissonconsulting/extras.svg?branch=master)](https://travis-ci.com/poissonconsulting/extras)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/poissonconsulting/extras?branch=master&svg=true)](https://ci.appveyor.com/project/poissonconsulting/extras)
[![Codecov test coverage](https://codecov.io/gh/poissonconsulting/extras/branch/master/graph/badge.svg)](https://codecov.io/gh/poissonconsulting/extras?branch=master)
[![License: MIT](https://img.shields.io/badge/License-MIT-green.svg)](https://opensource.org/licenses/MIT)
[![CRAN status](https://www.r-pkg.org/badges/version/extras)](https://cran.r-project.org/package=extras)
<!-- [![Tinyverse status](https://tinyverse.netlify.com/badge/extras)](https://CRAN.R-project.org/package=extras) -->
<!-- ![CRAN downloads](https://cranlogs.r-pkg.org/badges/extras) -->
<!-- badges: end -->

`extras` provides basic functions for Bayesian analyses
including implementations of simple S3 generics from the universals package; 
functions to summarise MCMC samples; and R translations of a handful of common functions 
from modeling languages such as BUGS/JAGS and C++.
    
## Installation

<!-- To install the latest release from [CRAN](https://cran.r-project.org) -->
```{r, eval=FALSE, echo=FALSE}
install.packages("extras")
```

To install the developmental version from [GitHub](https://github.com/poissonconsulting/extras)
```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("poissonconsulting/extras")
```

To install the latest developmental release from the Poisson drat [repository](https://github.com/poissonconsulting/drat)
```{r, eval=FALSE}
# install.packages("drat")
drat::addRepo("poissonconsulting")
install.packages("extras")
```

## Demonstration

### Simple Generics

The `extras` package provides simple S3 generics from the universals package like `dims()`.

```{r}
library(extras)

x <- 1:10
dim(x) # base R (returns NULL!)
dims(x)
```

as well as those to summarise MCMC samples like `svalue()` which gives the surprisal value
```{r}
set.seed(1)
x <- rnorm(100)
svalue(rnorm(100))
svalue(rnorm(100, mean = 1))
svalue(rnorm(100, mean = 2))
svalue(rnorm(100, mean = 3))
```

## R translations

The package also provides R translations of BUGS/JAGS functions such as `log<-` and `logit<-`.

```{r}
mu <- NULL
log(mu) <- 1
mu
```

And C++ functions such as `pow()`.
```{r}
2^3 # base R
pow(2, 3)
```

## Contribution

Please report any [issues](https://github.com/poissonconsulting/extras/issues).

[Pull requests](https://github.com/poissonconsulting/extras/pulls) are always welcome.

Please note that this project is released with a [Contributor Code of Conduct](https://github.com/poissonconsulting/extras/blob/master/CODE_OF_CONDUCT.md).
By contributing, you agree to abide by its terms.
