---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# extras

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![R build status](https://github.com/poissonconsulting/extras/workflows/R-CMD-check/badge.svg)](https://github.com/poissonconsulting/extras/actions)
[![Codecov test coverage](https://codecov.io/gh/poissonconsulting/extras/branch/master/graph/badge.svg)](https://codecov.io/gh/poissonconsulting/extras?branch=master)
[![License: MIT](https://img.shields.io/badge/License-MIT-green.svg)](https://opensource.org/licenses/MIT)
[![CRAN status](https://www.r-pkg.org/badges/version/extras)](https://cran.r-project.org/package=extras)
<!-- ![CRAN downloads](https://cranlogs.r-pkg.org/badges/extras) -->

<!-- badges: end -->

`extras` provides basic functions for Bayesian analyses; 
functions to summarise MCMC samples; and R translations of a handful of common functions 
from modeling languages such as BUGS/JAGS and C++ (STAN and TMB).
    
## Installation

<!-- To install the latest release from [CRAN](https://cran.r-project.org) -->
```{r, eval=FALSE, echo=FALSE}
install.packages("extras")
```

To install the developmental version from [GitHub](https://github.com/poissonconsulting/extras)
```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("poissonconsulting/extras")
```

## Demonstration

### Summarise MCMC Samples

The `extras` package provides functions to summarise MCMC samples like `svalue()` which gives the *surprisal value* (Greenland, 2019)

```{r}
set.seed(1)
x <- rnorm(100)
svalue(rnorm(100))
svalue(rnorm(100, mean = 1))
svalue(rnorm(100, mean = 2))
svalue(rnorm(100, mean = 3))
```

## R translations

The package also provides R translations of BUGS/JAGS functions such as `log<-` and `logit<-`.

```{r}
mu <- NULL
log(mu) <- 1
mu
```

And STAN (C++) functions such as `pow()`.

```{r}
2^3 # base R
pow(2, 3)
```

## References

Greenland, S. 2019. Valid P -Values Behave Exactly as They Should: Some Misleading Criticisms of P -Values and Their Resolution With S -Values. The American Statistician 73(sup1): 106–114. http://doi.org/10.1080/00031305.2018.1529625.

## Contribution

Please report any [issues](https://github.com/poissonconsulting/extras/issues).

[Pull requests](https://github.com/poissonconsulting/extras/pulls) are always welcome.

Please note that this project is released with a [Contributor Code of Conduct](https://github.com/poissonconsulting/extras/blob/master/CODE_OF_CONDUCT.md).
By contributing, you agree to abide by its terms.
