---
title: "Beta-binomial Deviance Residuals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Beta-binomial Deviance Residuals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(extras)
library(tidyr)
library(ggplot2)
library(viridis)
```


### Beta-binomial distribution

The beta-binomial distribution is useful when we wish to incorporate additional variation into the probability parameter of the binomial distribution, $p$.
It accomplishes this by drawing $p$ from a beta distribution.
The parameters of the beta-binomial are the number of trials, $n$, and the shape parameters of the beta distribution, $\alpha$ and $\beta$.

The likelihood of the beta-binomial can be described as follows:
$$P(x | n, \alpha, \beta) = \frac{\Gamma(n + 1)}{\Gamma(x + 1)\Gamma(n - x + 1)} 
\frac{\Gamma(x + \alpha) \Gamma(n - x + \beta)}{\Gamma(n + \alpha + \beta)} 
\frac{\Gamma(\alpha + \beta)}{\Gamma(\alpha) \Gamma(\beta)},$$
where $x$ is a natural number $≤ n$, $\alpha$ and $\beta$ are $>0$, and where $\Gamma()$ represents the gamma function.  

The log-likelihood is:  
$$\log(P(x | n, \alpha, \beta)) = \log(\Gamma(n + 1)) - \log(\Gamma(x + 1)) - \log(\Gamma(n - x + 1)) +
   \log(\Gamma(x + \alpha)) + \\ \log(\Gamma(n - x + \beta)) - \log(\Gamma(n + \alpha + \beta)) +
    \log(\Gamma(\alpha + \beta)) - \log(\Gamma(\alpha)) - \log(\Gamma(\beta)).$$

The expected value of the beta-binomial distribution is $\frac{n \alpha}{\alpha + \beta}$.
This is the same as the expected value of the binomial distribution ($n p$), with the $p$ replaced by the expected value of the beta distribution, $\frac{\alpha}{\alpha + \beta}$.  

A parameterization frequently used for the beta-binomial distribution uses this expected probability ($p$) as a parameter, with a dispersion parameter ($\theta$) that specifies the variance in the probability. 
Our particular parameterization of $\theta$ is slightly unconventional, but has useful properties when modelling.
When $\theta = 0$, the beta-binomial reverts to the binomial distribution.
When $\theta = 1$ and $p = 0.5$, the parameters of the beta distribution become $\alpha = 1$ and $\beta = 1$, which correspond to a uniform distribution for the beta-binomial probability parameter.

The relationships between $p$ and $\theta$ and $\alpha$ and $\beta$ are defined as follows:  
$$p = \frac{\alpha}{\alpha + \beta}$$
$$\theta = \frac{2}{\alpha + \beta}$$

$$\alpha = 2\cdot\frac{p}{\theta}$$
$$\beta = 2\cdot\frac{1 - p}{\theta}$$

### Deviance:  

Recall that the likelihood of a model is the probability of the data set given the model ($P(\text{data}|\text{model})$). 
The *deviance* of a model, $D$, is defined by: 
$$D(\text{model},\text{data}) = 2(\log(P(\text{data}|\text{saturated model})) - \log(P(\text{data}|\text{fitted model}))),$$
where the saturated model is the model that perfectly fits the data. 

The saturated model usually uses the value of $x$ in place of the mean parameter, thereby fitting through the data points.
For example, calculating the saturated log-likelihood of the normal distribution involves using the $x$ value(s) in place of the mean parameter $\mu$ in the log-likelihood.
Recall from above that the expected value of the binomial distribution is $n \cdot p$.
Since $n$ in this distribution is fixed, $p = \frac{x}{n}$ is used to calculate the saturated log-likelihood of the binomial distribution.

In the case of the beta-binomial distribution, there is some confusion about what should be used as the saturated model.
One suggested approach is to use the saturated log-likelihood of the binomial distribution, as described above, effectively setting the $\theta$ dispersion term to 0.
A second suggested approach is to use the same method as described for the binomial distribution (i.e., replacing $p$ with $\frac{x}{n}$), and holding the $\theta$ value constant.
Both of these approach have issues.

The problem with the first approach is that the comparison is always to the binomial model, whose likelihood profile has a higher peak than the beta-binomial likelihood profile (Fig. 1). 
With increases in $\theta$, the density of the likelihood profile is more dispersed (Fig. 1). 
This approach can result in large deviances even when the expected value of the fitted model is close to the saturated binomial model. 

```{r, echo = FALSE, fig.width = 7, fig.height = 4, fig.cap = "Fig. 1: Beta-binomial likelihood profile for $x = 1$ and $n = 5$, for different values of $\\theta$. $\\theta = 0$ corresponds to the binomial case."}

n_samp <- 1000
x <- 1
size <- 5
prob <- 0.3
theta <- c(0, 0.1, 0.5, 1)

prob_seq <- seq(0, 1, length.out = n_samp)

lik <- data.frame(prob_seq = prob_seq,
                  lik_0.0 = exp(log_lik_beta_binom(x, size, prob_seq, theta[1])),
                  lik_0.1 = exp(log_lik_beta_binom(x, size, prob_seq, theta[2])),
                  lik_0.5 = exp(log_lik_beta_binom(x, size, prob_seq, theta[3])),
                  lik_1.0 = exp(log_lik_beta_binom(x, size, prob_seq, theta[4])))

tb <- as_tibble(lik) |>
  pivot_longer(cols = c(lik_0.0, lik_0.1, lik_0.5, lik_1.0),
               names_to = "theta", names_prefix = "lik_",
               values_to = "lik")

ggplot(tb, aes(x = prob_seq, y = lik, colour = theta)) +
  geom_line() +
  scale_colour_viridis(discrete = TRUE) +
  xlab("p") +
  ylab("likelihood")
```

The problem with the second approach is that for some combinations of $p$ and $\theta$, the resultant deviance values are negative.
This is an issue for two reasons.
Firstly, it is nonsensical because by definition the saturated model should be the best possible fit to the data. 
And secondly, it is also computationally problematic, because calculating deviance residuals, as we will do in the next section, involves taking the square root of the deviance, which is not possible for a negative number.
Here is an example of a case where the deviance is negative.
Using $x = 1$, $n = 5$, $p = 0.3$, and $\theta = 0.5$, we calculate the saturated log-likelihood:

$$\alpha_{sat} = \frac{2 (\frac{x}{n})}{\theta} = \frac{2(0.3)}{0.5} = 0.8$$
$$\beta_{sat} = \frac{2 (1 - \frac{x}{n})}{\theta} = \frac{2 (1 - 0.3)}{0.5} = 3.2$$

$$ \log(P(x|n, p_{sat}, \theta_{sat}) = -1.355106,$$  

the fitted log-likelihood:  

$$\alpha_{fit} = \frac{2 (p)}{\theta} = \frac{2(0.3)}{0.5} = 1.2$$
$$\beta_{fit} = \frac{2 (1 - p)}{\theta} = \frac{2 (1 - 0.3}{0.5} = 2.8$$

$$\log(P(x|n, p_{fit}, \theta_{fit})) = -1.32999,$$  

and finally, the deviance:  

\begin{aligned}
D(n, p, \theta, x) &= 2(\log(P(x|n, p_{sat}, \theta_{sat})) - \log(P(x|n, p_{fit}, \theta_{fit}))) \\
&= 2(-1.355106 - -1.32999) \\
&= -0.050232.
\end{aligned}

The deviance in this case is smaller than expected. 
This is because the likelihood at the expected value, $p = \frac{x}{n}$ does not match the peak of the likelihood profile (Fig. 2). 

So what value of the $p$ parameter maximizes the likelihood?
The beta-binomial distribution has closed-form solutions only for certain values of $\alpha$ and $\beta$.
Thus, we cannot generalize a solution, and must instead search for the value of $p$ that maximizes the likelihood for each data point in order to calculate the saturated log-likelihood.
We will refer to the optimized $p$ value for the $i^{th}$ data point as $p_i^*$, and generally as $p^*$.
We use $p^*$ to calculate the saturated log-likelihood, which produces deviances that are (a) strictly positive, and (b) relative to the value of $\theta$.
The value of $p_i^*$ for the example case with $x = 1$, $n = 5$, and $\theta = 0.5$ is shown in Figure 2.

```{r, echo = FALSE, fig.width = 7, fig.height = 4, fig.cap = "Fig. 2: Beta-binomial likelihood profile for $x = 1$, $n = 5$, and $\\theta = 0.5$. The dashed vertical line shows the likelihood at the expected value of the beta-binomial distribution ($n \\cdot p$) where $p = \\frac{x}{n} = \\frac{1}{5} = 0.2$. As you can see, this is not the $p$ for which the likelihood is maximized. The solid vertical line shows the likelihood at its maximum point. In this case, $p^* = 0.26$."}

n_samp <- 1000
x <- 1
size <- 5
prob <- 0.3
theta <- 0.5

opt_beta_binom <- function(prob, x, size, theta) {
  -log_lik_beta_binom(x = x, size = size, prob = prob, theta = theta)
}

opt_p <- stats::optimize(opt_beta_binom, interval = c(0, 1), x = x,
                         size = size, theta = theta)$minimum

prob_seq <- seq(0, 1, length.out = n_samp)

lik <- data.frame(prob_seq = prob_seq,
                  lik = exp(log_lik_beta_binom(x, size, prob_seq, theta)))
p_df <- data.frame(val = c(x / size, opt_p),
                   probability = c("Expected value (x / n) = 0.20", "Optimized value (p*) = 0.26"))

ggplot(lik, aes(x = prob_seq, y = lik)) +
  geom_line(colour = viridis(1, begin = 0.66)) +
  geom_vline(data = p_df, mapping = aes(linetype = probability, xintercept = val)) +
  xlab("p") +
  ylab("likelihood")
```

### Deviance residuals

The unit deviance, $d_i$ refers to the deviance for the $i^{th}$ data point, $x_i$:  
$$d_i =  2(\log(P(x_i|n, p_i^*, \theta)) - \log(P(x_i|n, p, \theta))).$$
The *deviance residual* is the signed squared root of the unit deviance,  
$$r_i = \text{sign}(x_i - (n p_i^*)) \sqrt{d_i}.$$
The following plot shows histograms of deviance residuals for 10,000 randomly generated data points with $n = 50$ and various values of $p$ and $\theta$.

```{r, echo = FALSE, fig.height = 8, fig.width = 7, fig.cap = "Fig. 4: Histograms of deviance residuals for beta-binomial data simulated with $n = 50$, $p = \\{0.3, 0.5, 0.9 \\}$, and $\\theta = \\{0.0, 0.1, 0.5, 1.0 \\}$. The percentages within each histogram (i.e., each combination of $p$ and $\\theta$) sum to 100%."}

n_samp <- 10000
size <- 50
prob <- c(0.3, 0.5, 0.9)
theta <- c(0, 0.1, 0.5, 1)

res_mat <- matrix(NA, nrow = n_samp * length(theta), ncol = length(prob))

set.seed(101)
for (i in 1:length(prob)) {
  res <- NULL
  for (j in 1:length(theta)) {
    x <- ran_beta_binom(n_samp, size, prob[i], theta[j]) 
    res <- c(res, res_beta_binom(x, size, prob[i], theta[j]))
  }
  res_mat[, i] <- res
}

df <- as.data.frame(res_mat)
colnames(df) <- c("prob_0.3", "prob_0.5", "prob_0.9")
df$theta <- rep(theta, each = n_samp)

df <- df |> 
  pivot_longer(cols = c(prob_0.3, prob_0.5, prob_0.9), names_to = "prob", 
               names_prefix = "prob_", values_to = "res")

ggplot(df, aes(x = res)) +  
  geom_histogram(aes(y = stat(density) * 0.1), binwidth = 0.1) +
  scale_y_continuous(labels = scales::percent) +
  facet_grid(~prob ~theta, labeller = label_bquote(cols = theta: .(theta), rows = p: .(prob))) +
  xlab("Deviance residuals") + 
  ylab("Percent (%)")
```

Recall that $\theta = 0$ refers to the binomial case.
When $p = 0.5$ and $\theta > 0$, deviance residuals close to 0 become less common; as $\theta$ increases, the distribution of deviance residuals appears to become less continuous (Fig. 4).
When $p$ deviates from $0.5$ and $\theta > 0$, the distribution of the deviance residuals becomes increasingly skewed and discontinuous (Fig. 4).

Three main differences exist between $\theta = 0$ and $\theta > 0$ for cases when $p ≠ 0.5$:  
1. The majority of the density of the deviance residuals shifts to one side (the positive side for $p > 0.5$ and the negative side for $p < 0.5$).  
2. Gaps occur in the histogram of deviance residuals.  
3. Certain deviance residuals become more common, especially when $p$ is close to 0 or 1, and when $\theta$ is high.

The shift in the density of the deviance residuals is because the expected value for the saturated model which is used to calculate the sign of the residual ($n \cdot p^*$) becomes increasingly divergent from the actual expected value ($n \cdot p$) with increasing $\theta$.

The gaps in the distributions where $\theta > 0$ result from a change in the sign of the residual once a certain value of $x$ is reached.
At this point, the x-value is greater than the expected value, and the sign of the residual turns positive.
Figure 5 shows a plot of deviance residuals against $x$ values for the case where $n = 50$, $p = 0.3$, and $\theta = 0.1$; notice the jump into the positive domain when $x = 26$.

```{r, echo = FALSE, fig.height = 4, fig.width = 7, fig.cap = "Fig. 5: Deviance residuals plotted against the associated value of $x$. The $x$ values were simulated with $n = 50$, $p = 0.3$, and $\\theta = 0.1$."}
n_samp <- 10000
size <- 50
prob <- 0.3
theta <- 0.1

set.seed(101)
x <- ran_beta_binom(n_samp, size, prob, theta)
res <- res_beta_binom(x, size, prob, theta)

df <- data.frame(x = x,
                 res = res)

ggplot(df, aes(x = x, y = res)) +
  geom_point() +
  xlab("x") +
  ylab("Deviance Residual")
```

The residuals with huge peaks in density result from an increasingly skewed distribution of simulated data as $\theta$ gets bigger than 0.
When $p < 0.5$ the distribution of simulated data skews right, and when $p > 0.5$ the distribution of simulated data skews left.
For example, the peak in the bottom-right histogram in Figure 4 (where $p = 0.9$ and $\theta = 1.0$) is the deviance residual for $x = 50$.
The reason for the peak is that $x = 50$ is by far the most common value in the simulated data with these parameters (Fig. 6).

```{r, echo = FALSE, fig.height = 4, fig.width = 7, fig.cap = "Fig. 6: Histogram of 10,000 simulated data points ($x$) with $n = 50$, $p = 0.9$, and $\\theta = 1.0$."}
n_samp <- 10000
size <- 50
prob <- 0.9
theta <- 1.0

set.seed(101)
x <- ran_beta_binom(n_samp, size, prob, theta)
x <- as.data.frame(x)

ggplot(x, aes(x = x)) +
  geom_histogram(aes(y = (..count..)/sum(..count..)), binwidth = 0.1) +
  scale_y_continuous(labels = scales::percent) +
  ylab("Percent (%)")
```

Due to the skewed and discontinuous nature of distribution of the deviance residuals the goodness of fit of a beta binomial model cannot be assessed from the deviance residuals alone but must be evaluated using posterior predictive checking. 
